---
title: "Activity: Bird Species"
author: "Cael Walter"
date: "07/24/24"
output: html_notebook
---


# Set Up

```{r}
# Load `tidyverse` and `dcData` packages
library(tidyverse)
library(dcData)


# Load the `OrdwayBirds` data set into your RStudio environment from the `dcData` package
data("OrdwayBirds")




# Show a few commands used to inspect the data set a few different ways
head(OrdwayBirds)
summary(OrdwayBirds)
str(OrdwayBirds)



```

# Activity

When you're finished with this activity, you will have a graph that shows what time of year various species appear at the Katherine Ordway Natural History Study Area in Inver Grove Heights, MN.

### Step 0

Before we begin, the book gives us some instructions to select a few key variables and clean up the date formatting.  Use the commands given in the book, but be sure to study the functions used so you know what is happening.  In this case, we select `SpeciesName`, `Month`, and `Day`. Then a `mutate()` command converts each variable to character, and then converts the resulting "character string" to a number.

```{r}
# Get the data table & clean up dates (see Data Computing eBook)
OrdwayBirds <-
  OrdwayBirds %>%
  select( SpeciesName, Month, Day ) %>%
  mutate( Month = as.numeric(as.character(Month)), 
          Day = as.numeric(as.character(Day)))


# Inspect resulting data
head(OrdwayBirds)

```



### Step 1




```{r}
# unique species in original data (including mis-spellings)
unique_species_original <- unique(OrdwayBirds$SpeciesName)
print(unique_species_original)

#Therer are 275 unique species in the original data.

# unique species in the clean list
unique_species_clean <- unique(OrdwayBirds$SpeciesName)
print(unique_species_clean)

#There is still 275 unique species in the clean list.

```


### Step 2
  



```{r}
install.packages("dplyr")
library(dplyr)

# Load the datasets
data("OrdwayBirds")
data("OrdwaySpeciesNames")




# inspect OrdwaySpeciesNames
head(OrdwaySpeciesNames)
summary(OrdwaySpeciesNames)
str(OrdwaySpeciesNames)

# inner join
OrdwayBirds_joined <- inner_join(OrdwayBirds, OrdwaySpeciesNames, by = "SpeciesName")




# inspect the data after the join
head(OrdwayBirds_joined)
summary(OrdwayBirds_joined)
str(OrdwayBirds_joined)

```
The variable used for matching cases is SpeciesName.
Added variables: SpeciesNameCleaned, Sex, Age, BandNumber, TrapID, Weather, DataEntryPerson, BandingReport, Weight. 
### Step 3

We want to identify the top 5-6 bird species sighted at the Katherine Ordway Natural History Study Area.

```{r}
# count sightings by species in descending order
species_counts <- OrdwayBirds %>%
  group_by(SpeciesName) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

print(species_counts)

# a hint in the text book recommends displaying the top 10 species to choose our threshold
top_10_species <- head(species_counts, 10)
print(top_10_species)


# define a sightings threshold; produce a table with only birds belonging to major species
threshold <- 1000


major_species <- species_counts %>%
  filter(Count >= threshold) %>%
  pull(SpeciesName)

OrdwayBirds_major_species <- OrdwayBirds %>%
  filter(SpeciesName %in% major_species)


print(OrdwayBirds_major_species)



```


### Step 4

We now want to produce a month-by-month count for each of the major species, and then construct a chart to tell the story of what time of year the various species appear.

```{r}
# prepare glyph-ready data table called `ByMonth`
ByMonth <- OrdwayBirds_major_species %>%
  group_by(SpeciesName, Month) %>%
  summarise(count = n()) %>%
  ungroup()


print(ByMonth)


```

Now use the ggplot template below to generate the desired graph. (Remark you can also use esquisser or mplot to create a line chart instead of the barchart)



```{r}
# ggplot commands

ByMonth %>% 
  ggplot(aes(x = Month, y = count, color = SpeciesName)) + 
  geom_line()

ByMonth %>% ggplot(aes(x=Month,y=count ,fill=SpeciesName)) +
  geom_bar(stat='identity',position='stack', width=.9) 

```

According to the graph, we can answer the questions posed:

##### 1. Which species are present year-round?  
American Goldfinch: Present in most months, indicating year-round presence.
Black-capped Chickadee: Present in most months, indicating year-round presence.


##### 2. Which species are migratory, that is, primarily present in one or two seasons?
American Goldfinch: Primarily seen in months 9 and 10, indicating it may be migratory.
Slate-colored Junco: Primarily seen in months 4 and 10 indicating it may be migratory with peak presence in cooler months.




##### 3. What is the peak month for each major species? (bullet list)
American Goldfinch: Peak in month 10.
Black-capped Chickadee: Peak in month 11.
Slate-colored Junco: Peak in month 10.
Tree Sparrow: Peaks in month 3.



##### 4. Which major species are seen in good numbers for at least 6 months of the year?
Slate-colored Junco and American Goldfinch


```{r}

```

